---
title: "Reproducible Research: Peer Assessment 1"
author: 
output: html_document
keep_md: true
---

## Loading and preprocessing the data
Lets read the Activity data into a data frame.   
```{r}
activity = read.csv("activity.csv")
str(activity)
```



## What is mean total number of steps taken per day?
1. Lets calculate the total number of steps per day. 
```{r}
StepsPerDay = tapply(activity$steps, activity$date, sum, na.rm=T)
StepsPerDay
```

2. Histogram of the total number of steps taken each day.
```{r}
hist(StepsPerDay)
```

3. Lets calculate mean and median. 
```{r}
avg1 = mean(StepsPerDay)
med1 = median(StepsPerDay)
```
Mean of total number of steps taken per day is **`r avg1`**.

Median of total number of steps taken per day is **`r med1`**.



## What is the average daily activity pattern?

1. Here's a time series plot of 5-minute intervals and average number of steps taken, averaged across all days.
```{r}
AvgStepsPerInterval = tapply(activity$steps, activity$interval, mean, na.rm=T)
plot(names(AvgStepsPerInterval), AvgStepsPerInterval )
```

2. Lets check which interval has max. avg. number of steps. 
```{r}
MaxAvgStepsInterval = names(AvgStepsPerInterval)[which.max(AvgStepsPerInterval)]
```
The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps is **`r MaxAvgStepsInterval`**.



## Imputing missing values

1. Lets calculate total number of missing values. 
```{r}
RowsMissingValues=sum(is.na(activity))
```
Total number of missing values is **`r  RowsMissingValues`**.

2. Lets look at the boxplots to devise a simple imputation strategy for missing values. 
```{r}
boxplot(activity$steps ~ activity$date)
abline(lm(activity$steps ~ activity$date))
boxplot(activity$steps ~ activity$interval)
abline(lm(activity$steps ~ activity$interval))
```

* We will impute missing values with the median for the interval as steps per interval values seems to have less variability as compared to steps per date.


* Next we will calculate the median steps per interval as per above strategy.
```{r}
MedianStepsPerInterval = tapply(activity$steps, activity$interval, median, na.rm=T)
```

3. Now we will create a new dataset that is equal to the original dataset but with the missing data filled in. 
```{r}
activityNew = activity
for (i in seq(0,2355,by=5)) {
  activityNew$steps[is.na(activityNew$steps) & activityNew$interval==i] = MedianStepsPerInterval[names(MedianStepsPerInterval)==i]
}
```

4. Histogram of the total number of steps taken each day.
```{r}
StepsPerDay_New = tapply(activityNew$steps, activityNew$date, sum, na.rm=T)
StepsPerDay_New
hist(StepsPerDay_New)
```


* Now lets check how steps per day have changed due to imputation. 
```{r}
newMean = mean(StepsPerDay_New)
newMedian = median(StepsPerDay_New)
diffMean = mean(StepsPerDay_New) - mean(StepsPerDay)
diffMedian = median(StepsPerDay_New) - median(StepsPerDay)
```

The new mean and median total number of steps taken per day are **`r newMean`** and **`r newMedian`** respectively.

By imputing missing data, the mean has changed by **`r diffMean`** and the median has changed by **`r diffMedian`**.


* The mean and the standard deviation for total daily number of steps have changed as seen below: 
```{r}
summary(StepsPerDay)
summary(StepsPerDay_New)
sd(StepsPerDay)
sd(StepsPerDay_New)
```


## Are there differences in activity patterns between weekdays and weekends?

1. Lets create a new factor variable indicating whether a given date is a weekday or weekend. 
```{r}
activityNew$dayofweek = ifelse((weekdays(x = strptime(activityNew$date,"%Y-%m-%d")) %in% 
                              c("Saturday", "Sunday")), 
                            "weekend",
                            "weekday")
```

2. Here's the plot of 5-min interval and average number of steps taken, averaged across all weekdays or weekends. 
```{r}
require(ggplot2)
g = ggplot(data = activityNew, aes(interval, steps, group=dayofweek))
g + stat_summary(fun.y=mean, geom="line") + 
  facet_grid(dayofweek ~ .) +  
  xlab("Interval") +
  ylab("No. of steps") +
  ggtitle("Average no. of steps taken on weekday vs. weekend") 
```


--- End of report ---